\documentclass[brown, compress, mathserif]{beamer}

\mode<presentation>
{
  \usetheme{Singapore}
  \useinnertheme{rectangles}
  \setbeamertemplate{background canvas}[vertical shading][bottom=white,top=white]
  \setbeamertemplate{footline}[frame number]
  \setbeamertemplate{navigation symbols}{}
}


\AtBeginDocument{
  \pgfdeclareverticalshading{beamer@headfade}{\paperwidth}
  {
    color(0cm)=(bg);
    color(1.25cm)=(bg)
  }
}


\usepackage[english]{babel}
\usepackage{eurosym}
\usepackage{booktabs}
\usepackage{colortbl}

\input {highlight.sty}

\title{Survey of languages for data stream query processing}

\author{Lu\'{i}s Pureza \\ \texttt{pureza@student.dei.uc.pt}}

\institute[University of Coimbra]
{
  Department of Informatics Engineering\\
  University of Coimbra, Portugal
}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

\section{Introduction}

\begin{frame}{Two paradigms}
  \begin{itemize}
  \item Declarative (SQL based)
    \begin{itemize}
    \item Coral8 CCL
    \item StreamSQL
    \item Esper EPL
    \item UCLA's ESL
    \end{itemize}
  \item Procedural
    \begin{itemize}
    \item Apama MonitorScript
    \item Aleri's SPLASH
    \end{itemize}
  \end{itemize}
\end{frame}

\section{Declarative languages}

\begin{frame}{Declarative languages}
  \hlstd{}\hlkwa{INSERT INTO }\hlstd{PriceMicrosoft}\hspace*{\fill}\\
  \hlkwa{SELECT }\hlstd{price}\hspace*{\fill}\\
  \hlkwa{FROM }\hlstd{StockTrades }\hlkwa{WHERE }\hlstd{symbol }\hlsym{= }\hlstd{}\hlstr{'MSFT'}\hlstd{}\hspace*{\fill}\\
\end{frame}


\begin{frame}{On top of that\ldots}
  \begin{itemize}
  \item Windows
  \item Pattern matching
  \end{itemize}
\end{frame}

\subsection{Windows}

\begin{frame}{Windows}
  All kinds of windows:
  \begin{itemize}
  \item Fixed-size or temporal windows;
  \item Sliding or jumping windows;
  \item Partitioned windows.
  \end{itemize}
\end{frame}


\begin{frame}{Fixed-size sliding windows}
  \hlstd{}\hlkwa{INSERT INTO }\hlstd{AvgPriceMicrosoft}\hspace*{\fill}\\
  \hlkwa{SELECT }\hlstd{}\hlkwd{avg}\hlstd{}\hlsym{(}\hlstd{price}\hlsym{) }\hlstd{}\\
  \hlkwa{FROM }\hlstd{StockTrades }\hlkwa{KEEP }\hlnum{3 }\hlkwa{ROWS }\\
  \hlkwa{WHERE }\hlstd{symbol }\hlsym{= }\hlstd{}\hlstr{'MSFT'}\hlstd{}\hspace*{\fill}\\
  \mbox{}
  \begin{itemize}
  \item \hlkwa{KEEP} \hlnum{1} \hlkwa{ROW} is equivalent to \hlkwa{KEEP LAST ROW};
  \item \hlkwa{KEEP ALL} retains everything;
  \item \hlkwa{KEEP} \hlnum{10} \hlkwa{LARGEST DISTINCT BY} price
  \end{itemize}
\end{frame}


\begin{frame}{Fixed-size jumping windows}
  \hlstd{}\hlslc{{-}{-} Window is cleared every 3 rows}\hspace*{\fill}\\
  \hlstd{}\hlkwa{INSERT INTO }\hlstd{AvgPriceMicrosoft}\hspace*{\fill}\\
  \hlkwa{SELECT }\hlstd{}\hlkwd{avg}\hlstd{}\hlsym{(}\hlstd{price}\hlsym{) }\hlstd{}\\
  \hlkwa{FROM }\hlstd{StockTrades }\hlkwa{KEEP EVERY }\hlnum{3 }\hlkwa{ROWS }\\
  \hlkwa{WHERE }\hlstd{symbol }\hlsym{= }\hlstd{}\hlstr{'MSFT'}\hlstd{}\hspace*{\fill}\\
\end{frame}


\begin{frame}{Time-based sliding windows}
  \hlstd{}\hlkwa{INSERT INTO }\hlstd{AvgPriceMicrosoft}\hspace*{\fill}\\
  \hlkwa{SELECT }\hlstd{}\hlkwd{avg}\hlstd{}\hlsym{(}\hlstd{price}\hlsym{) }\hlstd{}\\
  \hlkwa{FROM }\hlstd{StockTrades }\hlkwa{KEEP }\hlnum{3 }\hlkwa{HOURS }\\
  \hlkwa{WHERE }\hlstd{symbol }\hlsym{= }\hlstd{}\hlstr{'MSFT'}\hlstd{}\hspace*{\fill}\\
\end{frame}


\begin{frame}{Time-based jumping windows}
  \hlstd{}\hlslc{{-}{-} Window is cleared after 3 hours}\hspace*{\fill}\\
  \hlstd{}\hlkwa{INSERT INTO }\hlstd{AvgPriceMicrosoft}\hspace*{\fill}\\
  \hlkwa{SELECT }\hlstd{}\hlkwd{avg}\hlstd{}\hlsym{(}\hlstd{price}\hlsym{) }\hlstd{}\\
  \hlkwa{FROM }\hlstd{StockTrades }\hlkwa{KEEP EVERY }\hlnum{3 }\hlkwa{HOURS }\\
  \hlkwa{WHERE }\hlstd{symbol }\hlsym{= }\hlstd{}\hlstr{'MSFT'}\hlstd{}\hspace*{\fill}\\
\end{frame}


\begin{frame}{Partitioned windows}
  \hlstd{}\hlslc{{-}{-} 10 rows per group, not 10 rows overall!}\hspace*{\fill}\\
  \hlstd{}\hlkwa{INSERT INTO }\hlstd{MaxPrices}\hspace*{\fill}\\
  \hlkwa{SELECT }\hlstd{}\hlkwd{max}\hlstd{}\hlsym{(}\hlstd{price}\hlsym{) }\hlstd{}\\
  \hlkwa{FROM }\hlstd{StockTrades }\hlkwa{KEEP }\hlnum{10 }\hlkwa{ROWS }\\
  \hlkwa{GROUP BY }\hlstd{symbol }\hlstd{}\hspace*{\fill}\\
\end{frame}


\begin{frame}[fragile]
  \frametitle{What about\ldots StreamSQL}
  \begin{itemize}
  \item Pretty much the same thing.
  \end{itemize}

\begin{verbatim}
SIZE size ADVANCE increment
  {TIME | TUPLES | ON field_identifier_w}
  [PARTITION BY field_identifier_p[,...]]
  [VALID ALWAYS]
  [OFFSET offset]
  [TIMEOUT timeout]
\end{verbatim}
\end{frame}


\begin{frame}{What about\ldots Esper EPL (1)}
  \begin{itemize}
  \item Mostly the same too.
    \bigskip\\
    \hlkwa{SELECT } \hlstd{tickDataFeed}, \hlkwd{stddev}(\hlstd{price})\\
    \hlkwa{FROM } \hlstd{StockTickEvent}\hlsym{(}\hlstd{symbol\hlsym{=}\hlstr{'IBM'})}.win\hlsym{:}length(\hlnum{10})\\
    \hlkwa{WHERE } \hlstd{volume} \hlsym{$>$} \hlnum{1000}\\
    \hlkwa{GROUP BY } tickDataFeed\\
    \hlkwa{HAVING } \hlkwd{stddev}(price) \hlsym{$>$} \hlnum{0.8}

  \end{itemize}
\end{frame}


\begin{frame}{What about\ldots Esper EPL (2)}
  \begin{itemize}
  \item Esper also supports \alert{istream} and \alert{rstream}
    \bigskip\\
    \hlkwa{SELECT RSTREAM } \hlsym{*}\\
    \hlkwa{FROM } \hlstd{StockTickEvent}\hlsym{(}\hlstd{symbol\hlsym{=}\hlstr{'IBM'})}.win\hlsym{:}time(\hlnum{10 sec})\\
  \end{itemize}
\end{frame}


\begin{frame}{What about\ldots ESL}
  \begin{footnotesize}
  \hlkwa{INSERT INTO} stdout\\
  \hlkwa{SELECT} IdemID, current\_time \\
  \hlkwa{FROM} ClosedAuction \hlkwa{AS} C,\\
  \hlkwa{ TABLE}(OpenAuction \hlkwa{OVER} (\hlkwa{RANGE} \hlnum{24} \hlkwa{HOUR PRECEDING} C)) \hlkwa{AS} O\\
  \hlkwa{WHERE} O.itemID = C.itemID;
  \end{footnotesize}
\end{frame}

\begin{frame}{Is this enough? (1)}
  \framesubtitle{Semantic windows}
  \begin{definition}
    Windows whose endpoints may depend on events rather than being fixed.
  \end{definition}

  \begin{example}
    How many people entered the room while I was there?
  \end{example}
\end{frame}

\begin{frame}{Is this enough? (2)}
  \framesubtitle{Future windows}
  \begin{center}
    \begin{example}
      The RFID reader must detect the product 10 seconds before or after it detects personnel.
    \end{example}
  \end{center}

  \begin{small}
    \hlkwa{SELECT} person.tagid\\
    \hlkwa{FROM} tag readings \hlkwa{AS} person\\
    \hlkwa{WHERE} person.tagtype = ’person’ \hlkwa{AND NOT EXISTS}\\
    \ (\hlkwa{SELECT} * \hlkwa{FROM} tag readings \hlkwa{AS} item\\
    \ \ \hlkwa{OVER} [\hlnum{10} \hlkwa{SECONDS PRECEDING AND FOLLOWING} person]\\
    \ \ \hlkwa{WHERE} item.tagtype = ’item’)
  \end{small}
\end{frame}

\subsection{Pattern matching}


\begin{frame}{Pattern matching}
  \hlstd{}\hlslc{{-}{-} Fast price increases}\hspace*{\fill}\\
  \hlkwa{INSERT INTO } StockGains\\
  \hlkwa{SELECT } First.tradeid, Second.tradeid\\
  \hlkwa{FROM }  StockTrades as First, StockTrades as Second\\
  \hlkwa{MATCHING }[\hlnum{5} seconds: First, Second]\\
  \hlkwa{ON    } First.symbol = Second.symbol\\
  \hlkwa{WHERE }(Second.price - First.price) $>$ \hlnum{1.0};
\end{frame}


\begin{frame}{Operators}
  \begin{itemize}
  \item Sequence (,)
  \item Negation (!)
  \item Conjunction (\&\&)
  \item Disjunction ($||$)
  \end{itemize}
\end{frame}


\begin{frame}[fragile]
  \frametitle{More examples}
  \begin{itemize}
  \item A followed by B and then not followed by any C within 10 seconds:
\begin{verbatim}
    [10 SECONDS: A, B, !C]
\end{verbatim}
  \item A B not preceded by an A in the previous 10 seconds and not followed by any C within the next 20 seconds:
\begin{verbatim}
    [20 SECONDS: [10 SECONDS: !A, B], !C]
\end{verbatim}
  \item An A or a B not followed by any C within the next minute:
\begin{verbatim}
    [1 MINUTE: A || B, !C]
\end{verbatim}
  \end{itemize}

\end{frame}

\begin{frame}{Is this enough?}
\framesubtitle{SASE+}
  \begin{center}
    \begin{example}
      Stock increases from 10 to 20 within 1 hour.
    \end{example}
  \end{center}
\hlkwa{PATTERN SEQ}(STOCK+ a[ ])\\
\hlkwa{WHERE} \hlkwd{skip\_till\_next\_match}(a[ ]!)\\
    \{\\
    \ \ \ \ [symbol] \hlkwa{AND}\\
    \ \ \ \   a[\hlnum{1}].price = \hlnum{10} \hlkwa{AND}\\
    \ \ \ \   a[i].price $>$ a[i-\hlnum{1}].price \hlkwa{AND}\\
    \ \ \ \   a[a.LEN].price = \hlnum{20}\\
    \}\\
\hlkwa{WITHIN} \hlnum{1} hour\\
\hlkwa{RETURN} a[\hlnum{1}].symbol, a[].price \hlstd{}\hlslc{{-}{-} Returns the list of prices!}\hspace*{\fill}\\
\end{frame}

\subsection{User defined aggregates}

\begin{frame}{User defined aggregates (UDAs)}
  \begin{itemize}
  \item SQL let's you define your own aggregates;
  \item How to implement UDAs for continuous queries?
  \end{itemize}
\end{frame}


\begin{frame}[fragile]
  \frametitle{UDAs in ESL}
  \begin{footnotesize}
    \begin{verbatim}
WINDOW AGGREGATE myavg(Next Real) : Real
{
    TABLE state(tsum Int, cnt Real);

    INITIALIZE : {
        INSERT INTO state VALUES (Next, 1);
    }

    ITERATE : {
        UPDATE state SET tsum=tsum+Next, cnt=cnt+1;
        INSERT INTO RETURN
            SELECT tsum/cnt FROM state;
    }

    EXPIRE: {
        UPDATE state SET cnt= cnt-1,
                 tsum = tsum - oldest()
    }
}
\end{verbatim}
  \end{footnotesize}
\end{frame}

\section{Procedural languages}

\subsection{Apama monitorscript}


\begin{frame}{Apama monitorscript}
  \begin{itemize}
  \item \textbf{monitors} are event processing ``agents'';
    \begin{itemize}
    \item Monitors can create listeners that await on events;
    \item When an event arrives, it processes it;
    \end{itemize}
  \item \textbf{route} creates new events that will be caught by other monitors;
  \item \textbf{actions} are similar to methods.
  \item Monitors can have state!
  \end{itemize}
\end{frame}


\begin{frame}[fragile]
  \frametitle{It looks like this}

  \begin{verbatim}
monitor ProcessMarket {
    action onload {
        Symbol sym;
        on all Symbol(): sym {
           processSymbol(sym);
           route SymbolAck(sym.id);
        }
    }

    action processSymbol(Symbol sym) {
       log "Received Symbol Event = " + sym;
    }
}
\end{verbatim}

\end{frame}

\subsection{Aleri's SPLASH}

\begin{frame}{Aleri's SPLASH}
  \begin{example}
    Suppose there is a stream of Trades data, with fields TradeId, Symbol
(e.g., MSFT), and Price. Suppose we'd like to stamp each IBM trade --- and only the IBM
trades --- with a sequence number.
  \end{example}
\end{frame}


\begin{frame}{SQL solution}
SequenceNumber =\\
 \hlkwa{ SELECT} max(T.TradeId) \hlkwa{as} TradeId, count(T.Symbol) \hlkwa{as} SeqNo\\
 \hlkwa{ FROM} Trades T\\
 \hlkwa{ WHERE} T.Symbol = \hlstr{'IBM'}\\
 \hlkwa{ GROUP BY} T.Symbol\\
\vspace{10 mm}
TaggedIBMTrades =\\
 \hlkwa{ SELECT} T.TradeId, T.Symbol, T.Price,\\
   (\hlkwa{CASE WHEN} T.Symbol = \hlstr{'IBM'} \hlkwa{THEN} S.SeqNo \hlkwa{ELSE} null \hlkwa{END}) \hlkwa{as} IBMNumber\\
 \hlkwa{ FROM} Trades T, SequenceNumber S\\
 \hlkwa{ WHERE} T.TradeId = S.TradeId
\end{frame}


\begin{frame}[fragile]
  \frametitle{SPLASH's solution}
\begin{verbatim}
int32 seqNo := 0;
if (Trades.Symbol = 'IBM') {
    output [
        TradeId = Trades.TradeId;
        Symbol = Trades.Symbol;
        IBMNumber = seqNo
    ];
    seqNo := seqNo + 1;
} else {
    output [
        TradeId = Trades.TradeId;
        Symbol = Trades.Symbol;
        IBMNumber = null
    ];
}

\end{verbatim}
\end{frame}

\begin{frame}{End!}
  \begin{center}
    Thank you!
  \end{center}
\end{frame}

\end{document}

